--- 
title: "Data Management and Reproducible Science"
author: "Claire S Teitelbaum"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
format:
  html:
    html-table-processing: none
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This book contains lessons and exercises for FANR 5950/8950, Data Management and Reproducible Science, Fall 2025."
---

# Course overview

This book contains lesson and exercises for FANR 5950/8950, Data Management and Reproducible Science, Fall 2025.

Other course materials, including the syllabus, are available on [eLC](https://uga.view.usg.edu/d2l/login).

## How to use this book

Throughout this book, you will find text interspersed with R code. Though not required, it is recommended that you follow along by copying R code from this book into a script on your computer. The code will appear like the one in the box below, and you can copy it by hovering over the box and clicking the clipboard icon in the top-right corner:

```{r, eval = F, echo = T}
# This is some R code
class_work <- "this"
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<!-- bookdown::render_book("index.Rmd") -->

## Some general guidance about coding and data management

* There is rarely one right way to write code. One way might be faster, and/or easier to read, and/or simpler, but there are always a *lot* of roads to get to the same destination.
* All programming languages are there to do what you tell them to do. They do that very well. What they do not do is what you want them to do! If your code is throwing errors or producing unexpected results, it's probably because it didn't understand what you wanted it to do. 
* To take the previous two points together: when in doubt, try taking another road and seeing if it leads you to the same place.
* Your system or code may be different than your collaborator's or friend's. That is usually not a problem, as long as they can understand how to find what they need and what to do with it when they do.

## Course schedule

Below is the current schedule for this course. This is subject to change as the semester progresses.

```{r, echo = F, message = F}
library(tidyverse)
library(kableExtra)
readxl::read_excel("../DataMgmtSchedule.xlsx") %>%
  mutate(Date = format(Date, "%A, %B %e, %Y")) %>%
  mutate_all(replace_na, "") %>%
  knitr::kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                font_size = 12) %>%
  column_spec(5, width_max = "5cm")
```
