--- 
title: "Data Management and Reproducible Science"
author: "Claire S Teitelbaum"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This book contains lessons and exercises for FANR 5950/8950, Data Management and Reproducible Science, Fall 2025."
---

# Course overview

This book contains lesson and exercises for FANR 5950/8950, Data Management and Reproducible Science, Fall 2025.

Other course materials, including the syllabus, are available on [eLC](https://uga.view.usg.edu/d2l/login).

## How to use this book

Throughout this book, you will find actions indicated by <span style="color:red">**bold red text**</span>. Most exercises will ask you to copy R code from this book into a script on your computer. The code will appear like the one in the box below, and you can copy it by hovering over the box and clicking the clipboard icon in the top-right corner:

```{r, eval = F, echo = T}
# This is some R code
class_work <- "this"
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<!-- bookdown::render_book("index.Rmd") -->
